The Effectiveness of Strategies to Contain SARS-CoV-2: Testing, Vaccinations, and NPIs

### 2022-03-10 src/paper/data/course_of_disease.tex
Go back to longer version, nature submission, update references. (#6)
Mostly done except for Smith reference, switching machines.

% Death and Recovery from ICU ----------------------------------------------------------
We take the survival probabilities and time to death and time until recovery
($\gamma_{stop\:icu\:a}$ and $\gamma_{dead,\:a}$) from intensive care from
\citet{Hinch2021}. They report time until death to have a mean of 11.74 days and a
standard deviation of 8.79 days. To match this we discretize that 41\% of individuals who
will die from Covid-19 do so after one day in intensive care, 22\% die after twelve days,
29\% after 20 days and 7\% after 32 days. Again, we rescale this for every age group
among those that will not survive. For survivors \cite{Hinch2020} reports a mean duration
of 18.8 days until recovery and a standard deviation of 12.21 days. We discretize this
such that of those who recover in intensive care, 22\% do so after one day, 30\% after 15
days, 28\% after 25 days and 18\% after 45 days.

@@

% vaccinations
The second route to immunity is through vaccination. Germany has mostly relied on the
Pfizer-BioNTech BNT162b2 and Oxford-AstraZeneca ChAdOx1-S vaccines with smaller shares of
the Moderna and Johnson\&Johnson vaccines \citep{Impfdashboard}. As \citet{Pritchard2021}
and \citet{Harris2021} find no difference in the effectiveness between the two most
common vaccines, we do not distinguish between vaccine types.

>>>>

% Death and Recovery from ICU ----------------------------------------------------------
We take the survival probabilities and time to death and time until recovery
($\gamma_{stop\:icu\:a}$ and $\gamma_{dead,\:a}$) from intensive care from
\citet{Hinch2021}. They report time until death to have a mean of 11.74 days and a
standard deviation of 8.79 days. To match this we discretize that 41\% of individuals who
will die from Covid-19 do so after one day in intensive care, 22\% die after twelve days,
29\% after 20 days and 7\% after 32 days. Again, we rescale this for every age group
among those that will not survive. For survivors \cite{Hinch2021} reports a mean duration
of 18.8 days until recovery and a standard deviation of 12.21 days. We discretize this
such that of those who recover in intensive care, 22\% do so after one day, 30\% after 15
days, 28\% after 25 days and 18\% after 45 days.

@@

% vaccinations
\label{r.2.c}
The second route to immunity is through vaccination. Germany has mostly relied on the
Pfizer-BioNTech BNT162b2 and Oxford-AstraZeneca ChAdOx1-S vaccines with smaller shares of
the Moderna and Johnson\&Johnson vaccines \citep{Impfdashboard}. As \citet{Pritchard2021}
and \citet{Harris2021} find no difference in the effectiveness between the two most
common vaccines, we do not distinguish between vaccine types.

### 2022-03-10 src/paper/report.tex
Mostly done except for Smith reference, switching machines.

In the model, susceptibility to contracting the SARS-CoV-2 virus is dependent on age
\citep{Davies2020}. A possible infection progresses as shown in
Figure~\ref{fig:model_disease_progression}. We differentiate between an initial period
of infection without being infectious or showing symptoms, being infectious
(presymptomatic or asymptomatic), showing symptoms, requiring intensive care, and
recovery or death \citep[similar to][]{Aleta2020}. The probabilities of transitioning
between these states depend on age; their duration is random and calibrated to medical
literature (for a detailed description see Supplementary
Material~\ref{sub:data_course_of_disease}). Conditional on the type of contact,
infectiousness is independent of age \citep{Jones2021}.

@@

A similar-sized effect---42\% in the decomposition---comes from rapid testing. Here, it
is crucial to differentiate between recorded cases and actual cases. Additional testing
means that additional infections will be recorded which would otherwise remain
undetected. Figure~\ref{fig:2021_scenarios_recorded} shows that this effect is large and
may persist for some time. Until late April, recorded cases are higher in the scenario
with rapid testing alone when compared to the setting where none of the three mechanisms
are turned on. The effect on total cases, however, is visible immediately in
Figure~\ref{fig:2021_scenarios_newly_infected}. Despite the fact that only 10\% of the
population performed weekly rapid tests in March on average, new infections on April~1
would have been reduced by 53\% relative to the scenario without vaccinations, rapid
tests, or seasonality.

@@

Having quantified the effects of various mechanisms, we now simulate hypothetical
scenarios comparing changes in NPIs and testing regimes. Two of the most contentious
NPIs concern schools and mandates to work from home. In many countries, schools switched
to remote instruction during the first wave, so did Germany. After the summer break,
they were operating at full capacity with increased hygiene measures, before being
closed again from mid-December onward. Some states started opening them gradually in
late February, but operation at normal capacity did not resume until the beginning of
June. Figure~\ref{fig:school_scenarios} shows the effects of different policies
regarding schools starting after Easter, at which point rapid tests had become widely
available. We estimate the realized scenario to have essentially the same effect as a
situation with closed schools. Under fully opened schools with mandatory tests, total
infections would have been 6\% higher; this number rises to 20\% without tests. These
effect sizes are broadly in line with empirical studies (e.g. \citet{Vlachos2021,
    Berger2021}, see Section~\ref{subsec:model_validation} for a comparison). In light of
the large negative effects school closures have on children and parents
\citep{Luijten2021, Melegari2021}---and in particular on those with low socio-economic
status---these results in conjunction with hindsight bias suggest that opening schools
combined with a testing strategy would have been beneficial. In other situations, and
particular when rapid test are not available at scale, trade-offs may well be different.

>>>>

In the model, susceptibility to contracting the SARS-CoV-2 virus is dependent on age
\citep{Davies2020}.\label{r.2.b} A possible infection progresses as shown in
Figure~\ref{fig:model_disease_progression}. We differentiate between an initial period
of infection without being infectious or showing symptoms, being infectious
(presymptomatic or asymptomatic), showing symptoms, requiring intensive care, and
recovery or death \citep[similar to][]{Aleta2020}. The probabilities of transitioning
between these states depend on age; their duration is random and calibrated to medical
literature (for a detailed description see Supplementary
Material~\ref{sub:data_course_of_disease}). Conditional on the type of contact,
infectiousness is independent of age \citep{Jones2021}.

@@

A similar-sized effect---42\% in the decomposition---comes from rapid testing. Here, it
is crucial to differentiate between recorded cases and actual cases. Additional testing
means that additional infections will be recorded which would otherwise remain
undetected. Figure~\ref{fig:2021_scenarios_recorded} shows that this effect is large and
may persist for some time. Until late April, recorded cases are higher in the scenario
with rapid testing alone when compared to the setting where none of the three mechanisms
are turned on. The effect on total cases, however, is visible immediately in
Figure~\ref{fig:2021_scenarios_newly_infected}. Despite the fact that only 10\% of the
population performed weekly rapid tests in March on average, new infections on April~1
would have been reduced by 53\% relative to the scenario without vaccinations, rapid
tests, or seasonality. \label{r.1.b}\Copy{r.1.b}{In
    Appendix~\ref{subsec:robustness_rapid_test_sensitivity}, we provide a detailed analysis
    of whether our results are robust regarding the sensitivity parameters we assume for
    rapid tests. Even if we take a very pessimistic stance, the effect is only reduced from
    42\% to 38\%.}

@@

Having quantified the effects of various mechanisms, we now simulate hypothetical
scenarios comparing changes in NPIs and testing regimes. Two of the most contentious
NPIs concern schools and mandates to work from home.\label{ref.2.a} In many countries,
schools switched to remote instruction during the first wave, so did Germany. After the
summer break, they were operating at full capacity with increased hygiene measures,
before being closed again from mid-December onward. Some states started opening them
gradually in late February, but operation at normal capacity did not resume until the
beginning of June. Figure~\ref{fig:school_scenarios} shows the effects of different
policies regarding schools starting after Easter, at which point rapid tests had become
widely available. We estimate the realized scenario to have essentially the same effect
as a situation with closed schools. Under fully opened schools with mandatory tests,
total infections would have been 6\% higher; this number rises to 20\% without tests.
These effect sizes are broadly in line with empirical studies (e.g. \citet{Vlachos2021,
    Berger2021}, see Section~\ref{subsec:model_validation} for a comparison). In light of
the large negative effects school closures have on children and parents
\citep{Luijten2021, Melegari2021}---and in particular on those with low socio-economic
status---these results in conjunction with hindsight bias suggest that opening schools
combined with a testing strategy would have been beneficial. In other situations, and
particular when rapid test are not available at scale, trade-offs may well be different.

### 2022-03-10 src/paper/results/robustness_to_sensitivity_assumptions.tex
Mostly done except for Smith reference, switching machines.

Our result that frequent and large scale rapid testing was responsible for a
large decrease in cases in the spring of 2021 in Germany seems to be at odds
with the general perception that rapid tests are unreliable. They also seem to
contradict recent studies showing that the actual sensitivity of rapid tests t
be lower than advertised (e.g.\cite{Scheiblauer2021}).

This is, however, not the case. It is important to distinguish studies that
analyze the sensitivity and specificity of rapid tests in isolation and studies
that analyze the aggregate effect of testing strategies within an
epidemiological model.

@@

Both profiles imply much lower sensitivities of rapid tests than used in our
main results, especially in the later stage of an infection. However, the main
results stay very robust. The original result was that rapid tests, seasonality
and vaccinations are responsible for 42\%, 43\% and 16\%, respectively. With the
average profile, the effect of rapid tests goes down to 41 \%. With the lower
envelope profile, which is an extremely unfavorable assumption, it becomes 38\%.

@@

The sensitivity of rapid tests depends on the viral load in the repiratory
tract. It is low at the beginning of an infection (especially before the onset
of infectiousness), high in the first few days of infectiousness and then
gradually decreasing towards the end of infectiousness.

@@

% what is available -> two step mapping
Unfortunately, such sensitivity profiles are not usually reported in studies. We
thus need to create them by combining two types of studies: 1. Studies that
report the viral load in terms of threshold cycle ($Ct$) values determined by
PCR test (e.g. \cite{Cosentino2022}, \cite{Ong2021}, \cite{Bonnet2022},
\cite{Jang2021} and \cite{Zuin2021}). 2. Studies that report the ensitivity of
rapid tests for different $Ct$ values (e.g. \cite{Scheiblauer2021},
\cite{Bruemmer2021}).

@@

% Properties of the different methods
In general, using discretized values can lead to an underestimation of peak
sensitivities and an overestimation of very low sensitivities. This is because
discretization is essentially a smoothing device. On the other hand, it hat the
advantage of simply working with published results, without introducing any
tuning parameters or other assumptions.

% resulting profiles.
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/results/figures/data/testing/sensitivity_params_with_different_methods}
    \caption{Rapid test sensitivity profiles}
    \floatfoot{\noindent \textit{Note:} The figure shows estimated sensitivities
    of rapid tests over the course of an infection. The x-axis shows days
    relative to the onset of infectiousness. The y-axis shows the estimated
    rapid test sensitivities. The grey lines are the raw sensitivity estimates
    obtained with different methods of dealing with discretized data. The blue
    line shows their average and the yellow line their lower envelope. The
    turquoise line are the sensitivities used for the main results of the paper.
    It is much higher than the new estimates in the later stage of an infection.
    Luckily this has only a very small effect on the overall effectiveness of
    rapid tests on the infection dynamics.}
    \label{fig:sensitivity_assumptions}
\end{figure}

\FloatBarrier

Figure \ref{fig:sensitivity_assumptions} shows that the updated sensitivity
estimates are lower than the ones used for our original results, especially
towards the end of an infection. However, the main results barely change.

@@

    \caption{Updated shapley decompositions}
    \floatfoot{\noindent \textit{Note:} The figure shows updated versions of the
    shapley decomposition in figure \ref{fig:2021_scenarios_decomposition}. The
    decomposition is based on 20 model runs with different random seeds. The
    share attributed to each channel is rounded to the next full percentage
    point to acknowledge the remaining sampling uncertainty.}
    \label{fig:robustness_to_sensitivity_assumptions}
\end{figure}

\FloatBarrier

%

>>>>

\Copy{r.1.b-appendix}{
    Our main results are based on rapid test sensitivities read from clinical trials. Recent studies showing that the actual sensitivity of rapid tests may be lower than that \citep[e.g.,][]{Scheiblauer2021}.

@@

    Both profiles imply lower sensitivities of rapid tests than used in our main
    results. This is especially true during the later stage of an infection. However,
    the main results stay very robust. The original result was that rapid tests,
    seasonality and vaccinations are responsible for 42\%, 43\% and 16\%, respectively.
    With the average profile, the effect of rapid tests decreases to 41 \%. With the
    lower envelope profile, which is an extremely unfavorable assumption, it becomes
    38\%.

@@

    The sensitivity of rapid tests depends on the viral load in the respiratory tract.
    It is low at the beginning of an infection (especially before the onset of
    infectiousness), high in the first few days of infectiousness and then gradually
    decreasing towards the end of infectiousness.

@@

    % what is available -> two step mapping
    Unfortunately, such sensitivity profiles are not usually reported in studies. We
    thus need to create them by combining two types of studies: 1. Studies that report
    the viral load in terms of threshold cycle ($Ct$) values determined by PCR test
    \citep[e.g.][]{Cosentino2022,Ong2021,Bonnet2022,Jang2021,Zuin2021}. 2. Studies that
    report the sensitivity of rapid tests for different $Ct$ values
    \citep[e.g.][]{Scheiblauer2021,Bruemmer2021}.

@@

    % Properties of the different methods
    In general, using discretized values can lead to an underestimation of peak
    sensitivities and an overestimation of very low sensitivities. This is because
    discretization is essentially a smoothing device. On the other hand, it has the
    advantage of simply working with published results, without introducing any tuning
    parameters or other assumptions.

    % resulting profiles.
    \begin{figure}
        \centering
        \includegraphics[width=0.9\textwidth]{figures/results/figures/data/testing/sensitivity_params_with_different_methods}
        \caption{Rapid test sensitivity profiles}
        \floatfoot{\noindent \textit{Note:} The figure shows estimated sensitivities of
            rapid tests over the course of an infection. The x-axis shows days relative
            to the onset of infectiousness. The y-axis shows the estimated rapid test
            sensitivities. The grey lines are the raw sensitivity estimates obtained
            with different methods of dealing with discretized data. The blue line shows
            their average and the yellow line their lower envelope. The turquoise line
            are the test sensitivities used for the main results of the paper.}
        \label{fig:sensitivity_assumptions}
    \end{figure}

    % \FloatBarrier

    Figure \ref{fig:sensitivity_assumptions} shows that the updated sensitivity
    estimates are lower than the ones used for our original results, especially towards
    the end of an infection. However, the main results barely change. This is due to the
    fact that the differences are most important towards the later stage of an
    infection. Uncovering an infection that was previously undetected at that stage does
    not have a large effect on infection dynamics.

@@

        \caption{Shapley decompositions for different values of rapid test sensitivity}
        \floatfoot{\noindent \textit{Note:} The figure shows updated versions of the
            Shapley decomposition in figure \ref{fig:2021_scenarios_decomposition}. The
            decompositions are based on 20 model runs with different random seeds. The
            share attributed to each channel is rounded to the next full percentage
            point to acknowledge the remaining sampling uncertainty.}
        \label{fig:robustness_to_sensitivity_assumptions}
    \end{figure}

}
\FloatBarrier

### 2022-03-10 src/rev1_reply_comments.tex
Mostly done except for Smith reference, switching machines.

    Many thanks for pointing out that we misclassified the Quidel SARS Sofia antigen
    FIA test; we removed the reference.

    Indeed, in the line of the additional evidence, the numbers we use for test
    sensitivity seem to be in the upper part of plausible values. On
    page~\pageref{r.1.a} of the paper, we cite the two references you pointed out (numbering of references as in this document):

@@

    We therefore followed
    your suggestion and ran additional analyses with more conservative assumptions on
    the sensitivity of rapid tests. We report on these in
    Appendix~\ref{subsec:robustness_rapid_test_sensitivity}. Our results are
    remarkably robust to this change. We write:

    The percentage of the decrease in spring 2021
    attributed to rapid tests decreases by only 4 percentage points even in the least
    favourable scenario. We describe our approach for calculating the adjusted
    sensitivities in the next answer. }

@@

              Other models also use the Mossong data: OpenABM-Covid-19 model
              (Hinch2020), Bicher2021.

              We only use the pre-pandemic diary data as the starting point. Our
              policies, which include non-mandated individual behavior adjustments, then
              reduce contacts taking the type of contact into account and for the most
              part not randomly but in very specific ways. This allows us to account for
              behavior reactions and policy effects that target different types of
              contacts differently. For example, household contacts continue to take
              place no matter the policy environment. Work from home does not keep
              random individuals at home but every individual that works has a
              time-constant value that determines when he or she will stay home. Which
              share of the population stays home is taken from Google mobility data. In
              the education sector we do not use the Mossong data but use completely
              mechanistic models. This gives us confidence that we can sufficiently
              accurately represent the contact patterns during the pandemic and account
              for changes induced by the pandemic.

>>>>

    Many thanks for pointing out that we misclassified the Quidel SARS Sofia antigen
    FIA test; we removed the reference.{\bf \color{red}DOUBLE-CHECK -- this is Smith2021}

    Indeed, in the line of the additional evidence, the numbers we use for test
    sensitivity seem to be in the upper part of plausible values. On
    page~\pageref{r.1.a} of the paper, we cite the two references you pointed out
    (numbering of references relates to this document):

@@

    We followed your suggestion and ran additional analyses with more conservative
    assumptions on the sensitivity of rapid tests. The results are very robust. We write
    in the main text on page~\pageref{r.1.b}

    \textit{\Paste{r.1.b}}

    The robustness is due to the fact that over the course of an infection, the largest differences between the parameters we use in our main analysis and those based on other studies are found during the later stages of an infection. Uncovering an infection that was previously undetected at that stage does
    not have a large effect on subsequent infection dynamics. Our whole analysis in
    Appendix~\ref{subsec:robustness_rapid_test_sensitivity} reads as follows:

    \textit{\Paste{r.1.b-appendix}}

@@

              Many thanks for this comment. Ideally, we would of course use real-time data. For example, several studies have used smartphone tracking data for this purpose. However, they are not detailed enough for our requirements, because they do not provide any information about \textit{who} is meeting with each other. These dimensions make up our contact networks, which are absolutely crucial.

              Furthermore, note that we only use the pre-pandemic diary data as the
              starting point. Contact restrictions and individual behavior adjustments
              reduce contacts. These are active throughout the period of analysis and
              they vary with the strictness of measure. In order to do so, it is crucial
              to take the type of contact into account; see for example our analysis of
              potential school closures on Page~\pageref{ref.2.a}. Closed schools mean
              that children have far fewer and parents have somewhat fewer contacts than
              at  and for the most part not randomly but in very specific ways. This
              allows us to account for behavior reactions and policy effects that target
              different types of contacts differently. For example, household contacts
              continue to take place no matter the policy environment. Work from home
              does not keep random individuals at home but every individual that works
              has a time-constant value that determines when he or she will stay home.
              Which share of the population stays home is taken from Google mobility
              data. In the education sector we do not use the Mossong data but use
              completely mechanistic models. This gives us confidence that we can
              sufficiently accurately represent the contact patterns during the pandemic
              and account for changes induced by the pandemic.
